---
title: "EaES 396 Spring 2021 Code"
author: "Gavin McNicol"
date: "2/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages:

```{r message = F}
library(tidyverse)
library(raster)
library(ncdf4)
library(sf)
library(rgdal)
library(rwunderground)
```

The goals of this independent study project are to:

  1. Evaluate recent large-scale carbon cycle and wetland extent products in Illinois, including:
      + USFWS National Wetlands Inventory vs. WAD2M (SWAMPS-GLWD) vs. GEIMS v3.5
      + 1 km Global Soil Respiration (Warner et al. 2019)
      + 25 km FLUXNET-CH4 Upscaling (tentative version) (McNicol et al. in prep.)
      + US Wetland Methane Model (Matthews et al. )
      + Carbon Tracker (...)
      
  2. Compare and contrast reference and created wetlands in Chicagoland in terms of:
      + Greenness
      + Phenology 
      + Land surface temperature 
      + Evapotranspiration
      + Hydrological datasets (e.g. connected USGS gages)
      
The independent study student this semester wants to gain experience in QGIS, not R.  So much of this code is preparing geotif datafiles to work with in QGIS. 

#### Re-write geospatial data (.nc) as .geotifs for use in QGIS

1. Evaluate recent large-scale carbon cycle and wetland extent products in Illinois.

**Capture 4 months per year Jan, Apr, Jul, Oct**

Start by creating a bounding box for Illinois:

```{r}
xmin <- -91.513079
ymin <- 36.970298
xmax <- -87.494756
ymax <- 42.508481

p1 = st_point(c(xmax, ymax))
p2 = st_point(c(xmin, ymin))
sfc = st_(p1, p2, crs = 4326)

bb <- st_bbox(sfc)
```

Get WAD2M (Wetalnd Extent and Dynamics Product) [Zhang et al. in review](10.5194/essd-2020-262Earth) :

```{r}
wad2m <- brick("/Volumes/LACIE SHARE/Stanford CH4/June 2020 Upscaling/Grids/WAD2M/gcp-ch4_wetlands_2000-2018_025deg.nc")
str(wad2m)
```

The dataset runs from 2000-2018 (n = 228). Create index to subset only J, A, J, O:

```{r}
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
all_months <- rep(months, 19)
month_subset <- c("Jan", "Apr", "Jul",  "Oct")
index <- all_months %in% month_subset
index_loc <- c(1:228)[index == T]
```

Apply index to subset months:

```{r}
wad2m_subset <- wad2m[[index_loc]]
```

Then crop to Illinois:

```{r}
wad2m_subset_IL <- crop(wad2m_subset, bb)
```

Try plotting the first year:

```{r}
plot(wad2m_subset_IL[[60:76]])
```

Looks good. Now save as raster stack:

```{r}
writeRaster(wad2m_subset_IL, "/Users/gavin/Documents/Personnel/Students/Independent Study/Mary Wrightsman/eaes-396-spring-2021/data/wad2m/wad2m_subset_IL.tiff", format = "GTiff")
```

